---
title: "Data Manipulation"
author: "Chhiring Lama"
date: "2024-09-19"
output: github_document

---

```{r setup, echo = FALSE, message = F}
library(tidyverse)
```

This document will sow to "manipulate" data.

Import the two datasets that we're going to manipulate

```{r import_datasets, message=FALSE}
litters_df <- read_csv("./Data/FAS_litters.csv", na = c("NA", ".", ""))

litters_df <- janitor::clean_names(litters_df)

pups_df <- read_csv("./Data/FAS_pups.csv", na = c("NA", ".", ""))

pups_df <- janitor::clean_names(pups_df)
                       
```

# `select`

Use `select()` to select variables

```{r}
select(litters_df, group, litter_number, gd18_weight)
```

```{r}
select(litters_df, group:gd18_weight)
```

```{r}
select(litters_df, -c(gd0_weight, gd18_weight)) ##deselects these columns
```

```{r}
select(litters_df, starts_with("gd")) ##selects columns whose name start with gd_
```

```{r}
select(litters_df, GROUP = group) ##selects group and renames them
rename(litters_df, GROUP = group) ##renames but keeps all columns as they are 

select(litters_df, litter_number, gd0_weight, everything()) ##first gives two columns and then everything else

relocate(litters_df, litter_number, gd0_weight) ##first gives two columns and then everything else
```

## Learning Assessment: 

In the pups data, select the columns containing litter number, sex, and PD ears.
```{r}
select(pups_df, litter_number, sex, pd_ears) 
select(pups_df, litter_number:pd_ears) ##as they are consecutive columns
```

# `filter`

```{r}
filter(litters_df, gd_of_birth == 20) ##runs a Boolean function and returns rows that match this criterion
```

```{r}
filter(litters_df, pups_born_alive >= 8) 
filter(litters_df, pups_born_alive != 18) 
```


```{r}
filter(litters_df, group == "Low8") 
filter(litters_df, group %in% c("Low8", "low7")) # either
```

```{r}
filter(litters_df, group %in% c("Low8", "low7"), pups_born_alive == 8) ##true for group and for pups born alive
```


```{r}
drop_na(litters_df) ## drops rows with na in any column
drop_na(litters_df, gd0_weight) ##drops rows with na in the column gd0_weight
```


## Learning Assessment
In the pups data:

Filter to include only pups with sex 1
Filter to include only pups with PD walk less than 11 and sex 2

### Solution
```{r}
filter(pups_df, sex == 1) 
filter(pups_df, pd_walk < 11, sex == 2) 
```

# `mutate`

```{r}
mutate(litters_df,
  wt_gain = gd18_weight - gd0_weight,
  group = str_to_lower(group)
)

mutate(litters_df, sq_pups = pups_born_alive^2)
```

## Learning Assessment: 
In the pups data:

Create a variable that subtracts 7 from PD pivot
Create a variable that is the sum of all the PD variables

### Solution
```{r}
mutate(pups_df, 
       subtracted_pd_pivot = pd_pivot - 7, 
       sum_pd_vars = sum(select(pups_df,starts_with("pd_"))))
```

# `arrange`

```{r}
arrange(litters_df, gd0_weight)

arrange(litters_df, desc(gd0_weight))

arrange(litters_df, pups_born_alive, gd0_weight)
```


# PIPING

If we use individual functions separately that is a lot of repetition of assigning each new function to the variable.

Option 1 (not to use):
```{r, eval=FALSE}
litters_df_raw = 
    read_csv("./data/FAS_litters.csv", na = c("NA", ".", ""))

litters_df_clean_names = janitor::clean_names(litters_df_raw)
litters_df_selected_cols = select(litters_df_clean_names, -pups_survive)
litters_df_with_vars = 
  mutate(
    litters_df_selected_cols, 
    wt_gain = gd18_weight - gd0_weight,
    group = str_to_lower(group))
litters_df_with_vars_without_missing = 
  drop_na(litters_df_with_vars, wt_gain)
litters_df_with_vars_without_missing
```

Option 2(not to use): can nest each code, however it is very difficult to read and troubleshoot when needed

```{r, eval=FALSE}
litters_df_clean = 
  drop_na(
    mutate(
      select(
        janitor::clean_names(
          read_csv("./data/FAS_litters.csv", na = c("NA", ".", ""))
          ), 
      -pups_survive
      ),
    wt_gain = gd18_weight - gd0_weight,
    group = str_to_lower(group)
    ),
  wt_gain
  )
```


This is when piping comes in

```{r}
litters_df_clean <- read_csv("./Data/FAS_litters.csv", na = c("NA", ".", "")) |>  
  janitor::clean_names() |>  
  select(-pups_born_alive) |> 
  filter(gd_of_birth == 20) |> 
  mutate(
    wt_gain = gd18_weight - gd0_weight,
    group = str_to_lower(group)) |> 
  drop_na()
  
```

If we want to run a linear regression on the weight gain: 
```{r}
read_csv("./Data/FAS_litters.csv", na = c("NA", ".", "")) |>  
  janitor::clean_names() |>  
  mutate(
    wt_gain = gd18_weight - gd0_weight) |> 
  lm(wt_gain ~ pups_born_alive, data = _) |>
  broom::tidy() ##shows in a cleaner dataframe format
```


## Learning Assessment: 

Write a chain of commands that:

loads the pups data
cleans the variable names
filters the data to include only pups with sex 1
removes the PD ears variable
creates a variable that indicates whether PD pivot is 7 or more days

### Solution
```{r}
pups_df_clean <- read_csv("./Data/FAS_pups.csv", na = c("NA", ".", "")) |>  
  janitor::clean_names() |>  
  filter(sex == 1) |> 
  select(-pd_ears) |> 
  mutate(
    pd_pivot_greater_7 = (pd_pivot >= 7))
```





