---
title: "Tidy Data"
output: github_document
---

I'm an R Markdown document! 

```{r setup, echo = FALSE, message = FALSE}
library(tidyverse)
library(readxl)
library(haven)
```

This document will show how to tidy data

## pivot longer
```{r}
pulse_df  <- read_sas("Data/public_pulse_data.sas7bdat") |> 
  janitor::clean_names()
```

This needs to go from wide to long format.

```{r}
pulse_tidy_df  <- read_sas("Data/public_pulse_data.sas7bdat") |> 
  janitor::clean_names() |> 
  pivot_longer(
    cols = bdi_score_bl:bdi_score_12m, 
    names_to = "visit", 
    values_to = "bdi_score",
    names_prefix = "bdi_score_"
  ) |> 
  mutate(visit = replace(visit, visit == "bl", "00m")) |> 
  relocate(id, visit)
  
```

One more example.

```{r}
litters_df <- read_csv("Data/FAS_litters.csv", na = c("NA", ".", "")) |> 
  janitor::clean_names() 
  
```

This needs to go in the long format where we combine weight in gds. 
```{r}
litters_tidy_df <- read_csv("Data/FAS_litters.csv", na = c("NA", ".", "")) |> 
  janitor::clean_names() |> 
  pivot_longer(
    cols = gd0_weight:gd18_weight, 
    names_to = "gestational_day", 
    values_to = "weight"
  ) |> 
  mutate(
    gestational_day = case_match(
      gestational_day, 
      "gd0_weight" ~ 0, 
      "gd18_weight" ~ 18))
```

## pivot wider

Let's make up an analysis result table

```{r}
analysis_df <-
  tibble(
    group = c("treatment", "treatment", "control", "control"), 
    time = c("pre", "post", "pre", "post"), 
    mean = c(4, 10, 4.5, 5)
  )
```

The data is tidy but the visualization of 2-way association between group and time is better in wide format

```{r}
tibble(
    group = c("treatment", "treatment", "control", "control"), 
    time = c("pre", "post", "pre", "post"), 
    mean = c(4, 10, 4.5, 5)
  ) |> 
  pivot_wider(
    names_from = time, 
    values_from = mean
  ) |> 
  knitr::kable()
  
```

## Bind Tables.

```{r}
fellowship_ring <- read_excel("Data/LotR_Words.xlsx", range = "B3:D6") |> 
  mutate(movie = "fellowship_ring")

two_towers <- read_excel("Data/LotR_Words.xlsx", range = "F3:H6") |> 
  mutate(movie = "two_towers")

return_king <- read_excel("Data/LotR_Words.xlsx", range = "J3:L6") |> 
  mutate(movie = "return_king")

lotr_df <- 
  bind_rows(fellowship_ring, two_towers, return_king) |> 
  janitor::clean_names() |> 
  pivot_longer(
    cols = female:male, 
    names_to = "sex", 
    values_to = "words"
  ) |> 
  relocate(movie) |> 
  mutate(race = str_to_lower(race))
```

Import `litters` dataset
```{r}
litter_df = 
  read_csv(
    "./data/FAS_litters.csv",
    na = c("NA", ".", "")) |>
  janitor::clean_names() |>
  separate(group, into = c("dose", "day_of_tx"), sep = 3) |>
  relocate(litter_number) |>
  mutate(
    wt_gain = gd18_weight - gd0_weight,
    dose = str_to_lower(dose))
```

Import `pups` dataset
```{r}
pup_df = 
  read_csv(
    "./data/FAS_pups.csv",
    na = c("NA", "", ".")) |>
  janitor::clean_names() |>
  mutate(
    sex = 
      case_match(
        sex, 
        1 ~ "male", 
        2 ~ "female"),
    sex = as.factor(sex)) 

```

Join the datasets!
```{r}
fas_df = 
  left_join(pup_df, litter_df, by = "litter_number") |> 
  relocate(litter_number, dose, day_of_tx)

fas_df
```

